# Tidy Data

Tidy data refers to having your data organized in a specific manner suitable for analysis. An obscene amount of time in data analysis is spent getting data into a tidy format. This chapter will walk through common problems and approaches in cleaning and tidying your data, that will make it easier for others to follow and easier for you to work across technologies and subject areas. Keeping your data in a tidy format for analysis will help because it is a fundamentally flexible way of working with data. Keeping scattered, lose data in spreadsheets is a sure way to cause confusion for yourself and others. 

## The Importance of Tables

If I could convince you of one thing, it would be the value in keeping your data in a table-based excel format. Again, this is true for data analysis purposes. If you are trying to do math or statistics, then the table based framework can sometimes fall short. But if you are concerned with working with data to draw conclusions, then tables are the way to go. 

## Cleaning vs Tidying

I hate cleaning, but love tidying. Unfortunately, as in with life, one must clean before one tidies. But let's start with some conceptual definitions. 

Cleaning refers to the process of scrubbing the data into a way that makes sense to you, the analyst. Oftentimes, and especially in public sector organizations, the data is not clean. Whether you are looking at the output of a Survey Monkey survey or a canned report that is run from the IT department, your data will come in all shapes and sizes. Cleaning data is the process of transforming data values into ones that make sense for the purposes of your analysis. 

Here is the first major departure from what you may have been taught about data analysis in Excel. When you get messy data *do not* change individual cell values (if you can at all help it). Recall from the introductory chapter the difference between cells and fields. Fields, as a reminder, are columns that represent one variable. Whenever possible, use data analysis tools to make changes to the entire field, rather than specific cells. Most data analysis software, outside of Excel, make it difficult or impossible to change individual cell values. This is important for several reasons, most of which we will get to in the next chapter on reproducibility. But for now, thinking in terms of fields, and making changes to entire fields, will save you *a lot* of work and  headache in the long run. Let's look at a sample dataset that may be similar to one you would encounter in real life. Here is a survey collected by a field manager of a local parks and recreation department on employment. 

```{r dirty-data, message=FALSE, warning=FALSE, echo=TRUE}
library(tidyverse)
library(DT)

sites <- tribble(
  ~"Employee", ~"Location", ~"Telecommute?", ~"Hire Date",
  "ron swanson", "Pawnee City Hall", "never", "Unknown",
  "Knope, Leslie", "Field Duty", "1 day/week", "2011-6-1",
  "Andy Dwyer", "sullivan street pit", "40 hours", "March 1, 2013",
  "Jerry Gergich", "City Hall", "never", "6/1/1985",
  "Garry Gergich", "City Hall", "never", "6/1/1985",
  "ben wyatt", "Partridge, Minnesota","" , "Jan. 1, 2010"
)

sites %>% datatable(
    extensions = 'Buttons', 
    options = list(dom = 'Bfrtip', 
                   buttons = 'excel',
                   searching = FALSE))
```

In this example, it would be trivial to go in to the Excel file and clean up the dates, names, and locations by hand. However, you could imagine this survey replicated for a department of forty employees. It quickly becomes unfeasible to make those edits by hand. When this is the case, there are functions in Excel and R that will make your life much easier. 

Here is annotated code for how I would go about cleaning this table in R. The friendly syntax of the `tidyverse` packages makes it easy to follow along, even if you aren't comfortable writing it yourself. You can accomplish all these similar transformations using [Get & Transform in Excel](https://support.microsoft.com/en-us/office/power-query-101-008b3f46-5b14-4f8b-9a07-d3da689091b5). 

```{r}
#creating a new table called "sites_cleaned", starting with the old table "sites"
sites_cleaned <- sites %>% 
  #switching the order of names that are backwards
  mutate(Employee = if_else(Employee == "Knope, Leslie", "Leslie Knope", Employee)) %>% 
  #seperate employee names into two columns
  separate(Employee, into = c("first_name", "last_name")) %>% 
  #renaming column names to standard format
  rename(location = Location,
         telecommute_hours =`Telecommute?`,
         hire_date = `Hire Date`) %>% 
  #changing first and last names and locations to Title Case
  mutate(first_name = str_to_title(first_name),
         last_name = str_to_title(last_name),
         location = str_to_title(location)) %>% 
  #coding location data to three categories, In Office, In Field, or Other
  mutate(location = case_when(
           str_detect(location, "City Hall") ~ "In Office",
           str_detect(location, "Field") ~ "In Field",
           str_detect(location, "Street") ~ "In Field",
           TRUE ~ "Other"),
  #coding hours to numeric
         telecommute_hours = case_when(
           telecommute_hours == "never" ~ 0,
           telecommute_hours == "1 day/week" ~ 8,
           telecommute_hours == "40 hours" ~ 40
         )
         )

#print to datatable
sites_cleaned%>% datatable(
    extensions = 'Buttons', 
    options = list(dom = 'Bfrtip', 
                   buttons = 'excel',
                   searching = FALSE))
  
  
```

This may seem like a lot of work for a small table. But as your data grows, it is much easier to operate on entire fields at a time. This is especially true in Get & Transform, which makes it very easy to transform and clean data using all the same steps as the `R` code above. As you can see, our table is now "cleaned" and organized consistently.




## Using lower level data

Let's introduce a slightly more complicated tidy data problem, using the same base data as before. 

```{r}
visits_retirements <- tribble(
  ~"Employee", ~"Phone Visits", ~"Phone Retirements", ~"Office Visits", ~"Office Retirements", ~"Online Visits", ~"Online Retirements",
  "Danielle", 6, 4, 11, 8, 23, 15,
  "Ramona", 11, 7, 5, 3, 18, 15,
  "Ross", 10, 8, 10, 7, 10, 9 
)

knitr::kable(visits_retirements, caption = "Visits to the Dept. of Retirement Services in a given month by employee and associated client retirements")
```

Hopefully you will see a similar pattern here. Now, there are three variables: Visit type, number of visits, and number of retirements. Again, this data works fine for presentation but could use tidying to ease in analysis. 

```{r}
visits_retirements %>%
  DT::datatable(
    extensions = 'Buttons', 
    options = list(dom = 'Bfrtip', 
                   buttons = 'excel',
                   searching = FALSE))
```

Try to tidy this in R or Excel Get and Transform. See this footnote^[powerquery hints] or look at the code if you need a hint. 


```{r}
visits_retirements_tidy <- visits_retirements %>%
  pivot_longer(cols = -Employee, 
               names_to = c("Visit Location", "Type"), 
               names_sep = " ")

print(visits_retirements_tidy)
```
In this case, we actually pivoted too far. It will probably be more useful to have the counts of visits and retirements in their own category. Keep in mind the scope of the observation-- It is perfectly valid for each to have their own column, as it is visits and retirements per month. 

```{r}
visits_retirements_tidy2 <- visits_retirements_tidy %>% 
  pivot_wider(id_cols = c(Employee, `Visit Location`, Type), names_from = Type, values_from = value)

print(visits_retirements_tidy2)
```

From here, it is easy to do calculations based on fields, rather than cells. For example, in R or Get and Transform, you could add the following:

```{r}
visits_pct <- visits_retirements_tidy2 %>% 
  mutate(pct_retirements = Retirements / Visits)

print(visits_pct)
```


And then, one of the most useful things you can do is develop formulas by grouping of rows. For example, you may want to know the total number of visits and retirements by retiree, regardless of visit location. That can be accomplished in a pivot table. 

```{r}

```


## How is this democratic?

 Thinking of your analysis in terms of tidy data is the first step on your way to democratic data analysis. 

## Conclusion


## Practices/Resources

  