# Pivot to Win It


Why did we go through all the trouble of tidying data in the last section? So we can pivot. Thinking in terms of pivots, aka fields and summaries, is the key to effective democratic data analysis

## Pivot Tables and the Meaning of Everything
 > *The Ultimate Answer to Life, The Universe and Everything is... **the pivot table***

-Douglas Adams/Corban Nemeth

Pivot tables are the world's most common, most helpful, and most underrated data analysis tool. If you understand the mechanics of the pivot table, you have everything you need to be a data analysis expert. PowerBI or Tableau interactive charts and graphs are just pivot tables in disguise. Understanding what is needed to make a pivot table work is the key to the wide world of data analysis. 

What is so important about pivot tables? First and foremost, pivot tables force you to think in terms of fields, not in terms of cells. In order for a pivot table to be effective, the data has to be organized in a table. And there is a right and a wrong way to putting data in a table. If your pivot table is not working properly, it is likely because your data isn't tidy. 

A pivot table groups data by field and allows the user to drag fields to the rows or columns of the pivot table. This is effective when each field is a variable (something that can change), and each row is a separate observation of some phenomena of interest. 

In short, pivot tables depend on **tidy data**. 

Recall our sample data from the last chapter:

```{r}
#We have already loaded the "tidyverse" library so we do not have to do it again
#we are editing the "visits" table already created by storing it in a new table pivot_visits
pivot_visits <- visits %>% 
  #using pivot_longer on every column except "employee" and setting the name of the new columns
  pivot_longer(-Employee, names_to = "Visit Type", values_to = "Number of Visits") 


knitr::kable(pivot_visits, caption = "Visits to the Dept. of Retirement Services in a given month")
```

Once the data has been tidied, it is easy to manipulate fields in pivot tables. 

```{r}
rpivotTable::rpivotTable(pivot_visits, width = "60%", height = "400px")
```

The key with pivot tables, such as the one above, is you are moving *fields* into the rows and columns of a new table. This is often referred to as "pivoting". You are then selecting a calculation to *summarize* the intersection of the *fields* that you drug into rows and columns. Above, you can click on the "Count" button to see the other ways this lightweight pivot table allows you to *summarize* your data. More powerful data analysis tools, such as PowerPivot in Excel, PowerBI, and several R packages allow you infinitely flexible formulas to define what happens at the intersection of fields in a pivot table. This is the core of all (most) data analysis. Think in terms of fields and calculated summaries and you are well on your way to becoming a democratic data master. 

It is  also easier to do a variety of calculations on the data now that is in a "tidy" format. For example, creating descriptive statistics across groups is very straightforward. This can also be accomplished in an Excel pivot table with the basic ["show values as" functions](https://support.microsoft.com/en-gb/office/show-different-calculations-in-pivottable-value-fields-014d2777-baaf-480b-a32b-98431f48bfec).

```{r}
desc_visits <- pivot_visits %>% 
  group_by(`Visit Type`) %>% 
  summarise(`Avg Visits` = mean(`Number of Visits`),
            `Total Visits` = sum(`Number of Visits`),
            `Std Dev of Visits` = sd(`Number of Visits`))

```

This is where pivot tables, and the way they force you to think about data, become *enormously* powerful. Enter **PowerPivot**, stage left. PowerPivot allows you to write very detailed and useful functions for the values that appear in the middle of a pivot table. Above, I used R to calculate the average number of visits, total visits, and standard deviation of visits by visit type. But what if I quickly wanted to view this by Employee instead? Or what if I wanted to group by both visit type and employee, and generate the same descriptions? R and PowerPivot make this very easy. I'm not going to go into detail on the mechanics and tutorials of PowerPivot, as it is extensively covered elsewhere-- see the Practices section of this guide. 

Here's the key-- Every fancy data dashboard, analysis technique, interactive visualization-- is a pivot table with a "calculated field" (formula for values in a pivot table). That's it. If you can think in terms of fields and calculated values based on those fields, PowerBI, PowerPivot, Tableau, Qlik, ggplot2, etc. will be second nature. The only thing that changes when crating a viz instead of a pivot table is that your fields that you drag into a pivot table become the axis of your chart, and your calculated field becomes a "mark" on your plot. If  you are interested more in a universal so-called "grammar of graphics" that is based on the principles outlined here, there is a wealth of resources available. (link to resources.)

## Advanced Data Summaries
Frequently, public administrators are interested in how things have changed over a period of time. Here's an example dataset that shows referrals to the same Dept. of Retirement Systems we have used in previous examples. 

## Conclusion
Thinking in terms of pivot tables is the core of democratic data analysis. This may not feel intuitive at first glance. But thinking this way forces the analyst to avoid the classic pitfalls of *un*democratic data analysis. Those pitfalls are the cell-based, hard-coded, pages and pages of tabs excel workbook nightmare that only the analyst themselves understands (and that's if they are lucky). Thinking in terms of the language of data analysis, emphasizing the use of fields and summaries, forces your analysis to be flexible, understandable, and reproducible. It is easy for an analyst who is also fluent in data analysis to pick up your work and immediately see how it operates because it is constructed using best practices and principles. It is also easy for someone else, or your future self, to write new calculations for data summaries, compare results, and use different fields.


## Practices

