--- 
title: "Doing Democratic Data Analysis"
author: "Corban Nemeth"
date: "`r Sys.Date()`"
output:
  bookdown::gitbook:
    includes:
      in_header: header.html
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "An opinionated guide to data analysis for public administrators and policy analysts"
always_allow_html: true
---

```{r setup, include=FALSE}
# Add a common class name for every chunks
knitr::opts_chunk$set(
  echo = TRUE)
```

```{r htmlTemp3, echo=FALSE, eval=TRUE}
codejs <- readr::read_lines("js/codefolding.js")
collapsejs <- readr::read_lines("js/collapse.js")
transitionjs <- readr::read_lines("js/transition.js")

htmlhead <- 
  paste('
<script>',
paste(transitionjs, collapse = "\n"),
'</script>
<script>',
paste(collapsejs, collapse = "\n"),
'</script>
<script>',
paste(codejs, collapse = "\n"),
'</script>
<style type="text/css">
.code-folding-btn { margin-bottom: 4px; }
.row { display: flex; }
.collapse { display: none; }
.in { display:block }
</style>
<script>
$(document).ready(function () {
  window.initializeCodeFolding("show" === "hide");
});
</script>
', sep = "\n")

readr::write_lines(htmlhead, path = "header.html")
```

# Preface {-}

I believe that data, *in the hands of public administrators and policy analysts*^[*Not IT departments*], has the power to transform the way government works. 

Big questions will, and should, be asked of big data— the role of government in regulating algorithmic bias, facial recognition, and consumer data privacy is a vital conversation. However, these topics should not detract or deter public administrators and policy analysts from leaning into **small data** for decision-making purposes. 

Public administrators and analysts who are data literate will be able to make and inform better decisions while avoiding the pitfalls posed by the latest technological trends. This book represents an opportunity for public administrators and policy analysts to join their subject matter expertise with foundation principles and practices of democratic data analysis— data analysis that is **transparent, relevant, and grounded in the context of ethical and effective governance**.

In this guide, I present an opinionated framework for data analysis in public sector organizations. By opinionated, I mean that I will teach you what I think is the right way to do things given my own experience as a public sector policy and data analyst. Your experience might differ-- and that's great. I hope that where you can use your experience in place of mine, you do to the fullest extent. With that in mind, it is often said that you have to know the rules to break them, so I will teach you the "rules" as I understand them. 

This is guide is not an excel how-to manual. My hope is that the principles and practices outlined here will allow you to explore whatever analysis tools you are interested in in a democratic manner. With that said, practical examples are given in Excel and in R.


